---
title: "Collaborative Project K-means and Hclusts"
format: html
editor: visual
---

## Download Packages

```{r}
#| echo: false
library(tidyverse)
library(readxl)
library(cluster)
library(factoextra)
```

You can add options to executable code like this

```{r}
d <- read_excel("41562_2024_2070_MOESM4_ESM.xlsx")
```

**Look at the dataset**

```{r}
glimpse(d)
skimr::skim(d)
```

**Select numeric variables only.** Clustering methods like K-means etc. need numeric variables

```{r}
cluster_d <- d |>
  select(where(is.numeric)) |>
  drop_na()
```

**Scale the data**

```{r}
d_scaled <- scale(cluster_d)
```

Scaling puts all variables on the same scale.

This is important because variables with large values could otherwise dominate the clustering.

For example, a variable measured in thousands would have more influence than a variable measured from 0 to 1.

**Choose the number of clusters**

```{r}
fviz_nbclust(d_scaled, kmeans, method = "wss")
fviz_nbclust(d_scaled, kmeans, method = "silhouette")
```

The WSS elbow plot looks for the point where adding more clusters stops improving the model much.

The silhouette plot shows how well-separated the clusters are.

**Run K-means clustering**

```{r}
set.seed(123)
kmeans_result <- kmeans(d_scaled, centers = 4, nstart = 25)
```

centers = 4 means I am asking R to find 4 clusters.

nstart = 25 tells R to try 25 different starting points and keep the best result

set.seed(123) makes the result reproducible.

**View the K-means results**

```{r}
kmeans_result
kmeans_result$cluster #tells you which cluster each row belongs to
```

**Visualizing K-means clusters**

```{r}
#shows how the observations are grouped into clusters
fviz_cluster(kmeans_result, data = d_scaled)
```

**Add K-means clusters back to the data**

```{r}
data_kmeans <- cluster_d |>
  mutate(kmeans_cluster = as.factor(kmeans_result$cluster))
```

This creates a new dataset with the cluster number added as a new column, so we can compare the original variables across clusters.

**Create a distance matrix for hierarchical clustering**

```{r}
dist_matrix <- dist(d_scaled)
```

Hierarchical clustering needs to know how far apart each observation is from every other observation, this line calculates those distances.

**Run hierarchical clustering**

```{r}
#ward.D2 is to create compact, balanced clusters
hclust_result <- hclust(dist_matrix, method = "ward.D2")
```

**Plot the dendrogram**

```{r}
plot(hclust_result, labels = FALSE, main = "Hierarchical Clustering Dendrogram")
rect.hclust(hclust_result, k = 4, border = 2:4) #draws boxes around 3 clusters
```

**Cut the trees into clusters**

```{r}
hclust_clusters <- cutree(hclust_result, k = 4) #each row is assigned to one hierarchical cluster
```

**Add hierarchical clusters back to the data**

```{r}
data_hclust <- cluster_d |>
  mutate(hclust_cluster = as.factor(hclust_clusters))
```

**Compare K-means and hierarchical clustering**

```{r}
table(
  Kmeans = data_kmeans$kmeans_cluster,
  Hclust = data_hclust$hclust_cluster
)
```

The table shows whether the two methods grouped observations similarly.

If many observations fall into the same combinations, the methods are somewhat agreeing.

**Summarize cluster characteristics**

```{r}
#calculates the average value of each variable within each K-means cluster
data_kmeans |>
  group_by(kmeans_cluster) |>
  summarise(across(everything(), mean, na.rm = TRUE))

#does the same thing for hierarchical clusters
data_hclust |>
  group_by(hclust_cluster) %>%
  summarise(across(everything(), mean, na.rm = TRUE))
```
